SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-05-13 scipy-1.0/paper.tex
minor rewording / style fixes

% NOTE: some usage stats below are from https://libraries.io/pypi/scipy
% as well as github metrics; normally one doesn't put citations directly
% in abstract though (depends on field / journal)
\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard for leveraging scientific algorithms
in the Python programming language, with more than 600
unique code contributors, millions of downloads per year, 161 dependent
packages, and 28700 dependent repositories. This includes usage of SciPy
in almost half of all machine learning projects on GitHub, and usage by
high profile projects including LIGO gravitational wave analysis and creation
of the first-ever image of a black hole (M87).
The library includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we provide an overview of the capabilities and development practices of the
SciPy library and highlight some recent technical developments.
\end{abstract}
\begin{document}

@@

SciPy is a library of numerical routines for the Python programming
language that provides fundamental building blocks for modeling and
solving scientific problems. 
SciPy includes algorithms for optimization, integration, interpolation,
eigenvalue problems, algebraic equations, differential equations, and many other
classes of problems; it also provides
specialized data structures, such as sparse arrays
and $k$-dimensional trees. SciPy is built on top of 
NumPy\cite{vanderwalt2011numpy,Oliphant-2015},
which provides array data structures and related fast numerical routines, and
SciPy is itself the foundation upon which higher level scientific libraries,
including scikit-learn\cite{pedregosa2011scikit}
and scikit-image\cite{vanderwalt2014scikit}, are built. 
SciPy is relied upon by scientists, engineers, and others 
around the world. For example, published 
scripts\cite{alex_nitz_2018_1596771, LIGO-open}
used in the analysis of gravitational waves
\cite{PhysRevD.93.122003, abbott2017gw170817} 
import several submodules of SciPy, and the M87 black
hole imaging project directly cites SciPy\cite{2019ApJ875L3E}.

@@

\subsection*{SciPy Today}
At the time of writing, the SciPy library consists of nearly 
600,000 lines of code organized in 16 subpackages. 
Over 85,000 GitHub repositories and almost 5000 packages depend
on SciPy. Some of the major
feature highlights from the three years preceding
SciPy 1.0 are discussed in Section \ref{sec:technical_improvements}, 
and milestones in its history are highlighted in Figure~\ref{fig:timeline}.
% At the time of writing, SciPy produces a new release
% approximately every six months, and some of the major
% feature highlights from the three years preceding
% SciPy 1.0 are discussed in a technical improvements
% section below. Some longer-term milestone events are
% highlighted in Figure~\ref{fig:timeline}.

@@

The \texttt{scipy.stats} module provides more than 100 probability
distributions and a framework to implement additional distributions. Given an
expression for a probability distribution function (pdf), the framework will
calculate moments, cumulative distribution function (cdf), differential
entropy, and more. Those properties are calculated based on generic code, e.g.,
numerical integration of the pdf to obtain the cdf.  Where possible, the
generic code is replaced by explicit formulas or more accurate or efficient
implementations. For example, most distributions have an explicit formula for
the cdf, so it is used instead of integrating the pdf. The distributions
framework also allows the user to draw random variates from the distribution.
Drawing random variates now accepts a \texttt{random\_state} parameter, which
is either a \texttt{numpy.random.RandomState} object or an integer used to
generate such an object, to provide fine-grained user-level control over the
random state of a workflow.  Currently about 80 continuous univariate
distributions, 12 discrete univariate distributions and 10 multivariate
distributions are implemented.  Key recent distributions added to
\texttt{scipy.stats} include the histogram-based distribution in
\texttt{scipy.stats.rv\_histogram} and the multinomial distribution in
\texttt{scipy.stats.multinomial} (used, for example, in natural language
processing\cite{Griffiths5228}). \texttt{multinomial} is the first and so far
the only multivariate discrete distribution available in SciPy.

@@

\subsection*{Downstream projects}

The scientific Python ecosystem includes many examples
of domain-specific software libraries that are built 
on top of SciPy features. The work done on these libraries
often turns out to be beneficial to the whole ecosystem
being SciPy the catalyst of this cross-fertilisation process. 
For example, there are common contributors to the SciPy and 
Astropy core libraries\cite{astropy-2018}, and what works 
well for one of the codebases, infrastructures, or communities 
is often transferred in some form to the other. At the codebase 
level, the \texttt{binned\_statistic} functionality
is one such cross-project contribution---it was initially
developed in an Astropy-affiliated package
and then placed in SciPy afterwards.

@@

% The Discussion should be succinct and must not contain subheadings.

SciPy has a strong developer community and a massive user base. GitHub traffic
metrics report roughly 20,000 unique visitors to the source website between May
14 and May 27, 2018, with 721 unique copies (``clones") of the codebase over
that roughly two-week period. The developer community at that time consisted of 610 unique
contributors of source code, with $>19,000$ commits accepted into the codebase
(GitHub page data).

From the user side, there were 13,096,468 downloads of SciPy from the Python
Packaging Index (PyPI) during the year 2017\cite{pypinfo}, which is a lower
bound on the number of total downloads by users given that downloading from
PyPI is only one of several popular methods to install SciPy.  The SciPy
website\cite{SciPylib}, which has been the default citation in the absence of a
peer-reviewed paper, has been cited $>3000$ times. Some of the most prominent
usages of or demonstrations of credibility for SciPy include the LIGO / Virgo
scientific collaboration that lead to the observation of gravitational waves
\cite{PhysRevLett.116.061102}, the fact that SciPy is shipped directly with
MacOS and in the Intel Distribution for Python\cite{intel-python}, and that SciPy is used
by 47 \% of all machine learning projects on GitHub\cite{octoverse-scipy}.

>>>>

% NOTE: some usage stats below are from https://libraries.io/pypi/scipy
% as well as github metrics; normally one doesn't put citations directly
% in abstract though (depends on field / journal)
\begin{abstract}
SciPy is an open source scientific computing library for the Python programming language.
SciPy 1.0 was released in late 2017, about 16 years after the original
version 0.1 release. SciPy
has become a \emph{de facto} standard for leveraging scientific algorithms
in the Python programming language, with more than 600
unique code contributors, millions of downloads per year, 161 dependent
packages, and 28700 dependent repositories. This includes usage of SciPy
in almost half of all machine learning projects on GitHub.
The library includes functionality spanning clustering, Fourier transforms,
integration, interpolation, file I/O, linear algebra, image processing,
orthogonal distance regression, minimization algorithms, signal processing,
sparse matrix handling, computational geometry, and statistics. In this
work, we provide an overview of the capabilities and development practices of the
SciPy library and highlight some recent technical developments.
\end{abstract}
\begin{document}

@@

SciPy is a library of numerical routines for the Python programming
language that provides fundamental building blocks for modeling and
solving scientific problems. 
SciPy includes algorithms for optimization, integration, interpolation,
eigenvalue problems, algebraic equations, differential equations, and many other
classes of problems; it also provides
specialized data structures, such as sparse arrays
and $k$-dimensional trees. SciPy is built on top of 
NumPy\cite{vanderwalt2011numpy,Oliphant-2015},
which provides array data structures and related fast numerical routines, and
SciPy is itself the foundation upon which higher level scientific libraries,
including scikit-learn\cite{pedregosa2011scikit}
and scikit-image\cite{vanderwalt2014scikit}, are built. 
SciPy is relied upon by scientists, engineers, and others 
around the world. For example, published 
scripts\cite{alex_nitz_2018_1596771, LIGO-open}
used in the analysis of gravitational waves
\cite{PhysRevD.93.122003, abbott2017gw170817} 
import several submodules of SciPy.

@@

\subsection*{SciPy Today}
At the time of writing, the SciPy library consists of nearly 
600,000 lines of code organized in 16 subpackages. 
Over 85,000 GitHub repositories and almost 5000 packages depend
on SciPy. SciPy's GitHub repository attracts approximately 20,000 unique 
visitors every two weeks, and more than 100 people contribute to a 
new release every six months. Some of the major
feature highlights from the three years preceding
SciPy 1.0 are discussed in Section \ref{sec:technical_improvements}, 
and milestones in its history are highlighted in Figure~\ref{fig:timeline}.
% At the time of writing, SciPy produces a new release
% approximately every six months, and some of the major
% feature highlights from the three years preceding
% SciPy 1.0 are discussed in a technical improvements
% section below. Some longer-term milestone events are
% highlighted in Figure~\ref{fig:timeline}.

@@

The \texttt{scipy.stats} module contains more than 100 probability
distributions: 96 continuous and 13 discrete univariate distributions
and 10 multivariate distributions. The implementation relies on a
consistent framework that provides methods to sample random variates,
to evaluate the cumulative distribution function (cdf) and the probability
density function (pdf) and to fit parameters for every distribution. 
Generally, the methods rely on specific implementations for each
distribution such as a closed form expression of the cdf or a sampling 
algorithm, if available. Otherwise, default methods are used
based on generic code, e.g., numerical integration of the pdf to
obtain the cdf. 
Key recent distributions added to \texttt{scipy.stats} include the
histogram-based distribution in \texttt{scipy.stats.rv\_histogram}
and the multinomial distribution in \texttt{scipy.stats.multinomial}
(used, for example, in natural language processing, see
\cite{Griffiths5228}).

@@

\subsection*{Community beyond the SciPy library}

The scientific Python ecosystem includes many examples
of domain-specific software libraries building on top
of SciPy features, and then returning to the base SciPy library
to suggest and even implement improvements to enable
more effective science applications. For example, there
are common contributors to the SciPy and astropy core
libraries\cite{astropy-2018}, and what works well for 
one of the code bases, infrastructures, or communities 
is often transferred in some form to the other. At the code
base level, the \texttt{binned\_statistic} functionality
is one such cross-project contribution---it was initially
developed in an Astropy-affiliated package
and then placed in SciPy after the fact.

@@

\fixme{The Discussion should be succinct and must not contain subheadings.}

\subsection*{Impact now}

% While Paul was one of the early adopters at the national laboratories,
% Python's presence is now ubiquitous.

SciPy has a strong developer community and a massive user base. GitHub traffic
metrics report roughly 20,000 unique visitors to the source website between May
14 and May 27, 2018, with 721 unique copies (``clones") of the code base over
that roughly two-week period. The developer community consists of 610 unique
contributors of source code, with $>19,000$ commits accepted into the code base
(GitHub page data).

From the user side, there were 13,096,468 downloads of SciPy from the Python
Packaging Index (PyPI) during the year 2017\cite{pypinfo}, which is a lower
bound on the number of total downloads by users given that downloading from
PyPI is only one of several popular methods to install SciPy.  The SciPy
website\cite{SciPylib}, which has been the default citation in the absence of a
peer-reviewed paper, has been cited $>3000$ times. Some of the most prominent
usages of or demonstrations of credibility for SciPy include the LIGO / Virgo
scientific collaboration that lead to the observation of gravitational waves
\cite{PhysRevLett.116.061102}, and the fact that SciPy is shipped directly with
MacOS and in the Intel Distribution for Python\cite{intel-python}. SciPy is used
by 47 \% of all machine learning projects on GitHub\cite{octoverse-scipy}.

\subsection*{Future Work}

### 2019-05-13 scipy-1.0/subpackages.tex
minor rewording / style fixes

    \texttt{scipy.sparse.csgraph} provides a collections of graph algorithms
    for which the graph is represented using a sparse matrix.  Algorithms
    include connected components, shortest path, minimum spanning tree
    and more.
\item[\texttt{spatial}]
    This subpackage provides spatial data structures and algorithms,
    including the $k$-d tree, Delaunay triangulation, convex hulls and Voronoi
    diagrams.  The subpackage \texttt{scipy.spatial.distance} provides
    a large collection of distance functions, along with functions for
    computing the distance between all pairs of vectors in a given collection
    of points or between all pairs from two collections of points.
\item[\texttt{special}]
    The name comes the class of functions traditionally known as ``special
    functions'', but over time, the subpackage has grown to include functions
    beyond the classical special functions.  A more appropriate characterization
    of this subpackage is simply ``useful functions''.
    It includes a large collection of the classical special functions
    such as Airy, Bessel, etc.; orthogonal families of polynomials;
    the Gamma function, and functions related to it;
    functions for computing the PDF, CDF and quantile function for several
    probability distributions;
    information theory functions;
    combinatorial functions \texttt{comb} and \texttt{factorial};
    and more.
\item[\texttt{stats}]
    The \texttt{stats} subpackage provides a large collection of continuous
    and discrete probability distributions, each with methods to compute
    the PDF or PMF, CDF, moments and other statistics, generation of random
    variates, and more;
    statistical tests, including Pearson's correlation, Spearman's rank-order
    correlations, Kendall's tau, chi-squared test and its generalization as the
    Cressie-Read power divergence, contingency table tests including Fisher's
    exact test and Mood's median test, and many more;
    and assorted transformations and statistics of data.
\end{description}

>>>>

    \texttt{scipy.sparse.csgraph} provides a collections of graph algorithms
    for which the graph is represented using a sparse matrix.  Algorithms
    include connected components, shortest path, minimum spanning tree
    and more.
\item[\texttt{spatial}]
    This subpackage provides spatial data structures and algorithms,
    including the $k$-d tree, Delaunay triangulation, convex hulls and Voronoi
    diagrams.  The subpackage \texttt{scipy.spatial.distance} provides
    a large collection of distance functions, along with functions for
    computing the distance between all pairs of vectors in a given collection
    of points or between all pairs from two collections of points.
\item[\texttt{special}]
    The name comes the class of functions traditionally known as ``special
    functions'', but over time, the subpackage has grown to include functions
    beyond the classical special functions.  A more appropriate characterization
    of this subpackage is simply ``useful functions''.
    It includes a large collection of the classical special functions
    such as Airy, Bessel, etc.; orthogonal families of polynomials;
    the Gamma function, and functions related to it;
    functions for computing the PDF, CDF and quantile function for several
    probability distributions;
    information theory functions;
    combinatorial functions \texttt{comb} and \texttt{factorial};
    and more.
\item[\texttt{stats}]
    The \texttt{stats} subpackage provides a large collection of continuous
    and discrete probability distributions, each with methods to compute
    the PDF or PMF, CDF, moments and other statistics, generation of random
    variates, and more;
    statistical tests, including tests on equality of means/medians/variance
    (such as the t-test) and tests whether a sample is drawn from a
    certain distribution (such as the Kolmogorov-Smirnov test);
    measures of correlation, including Pearson's $r$, Kendall's $\tau$, and
    Spearman's $\rho$ coefficients;
    descriptive statistics including trimmed values;
    kernel density estimation;
    and transformations of data such as the Box-Cox power transformation.
\end{description}
